%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDICES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Ground Truth Annotation Dataset}
\label{appendix:ground-truth}

This appendix provides details on the ground truth annotation dataset used for system evaluation. The complete dataset comprises 30 YouTube videos with 503 annotated claims across three thematic categories.

%-------------------------------------------------------
\subsection{Video Corpus Summary}
%-------------------------------------------------------

Table~\ref{tab:video-corpus} presents the evaluation video corpus organized by category and content type.

\small
\begin{longtable}{p{6.5cm}p{2cm}p{3cm}r}
\caption{Evaluation video corpus by category and content type} \label{tab:video-corpus} \\
\toprule
\textbf{Video Title} & \textbf{Category} & \textbf{Content Type} & \textbf{Claims} \\
\midrule
\endfirsthead
\toprule
\textbf{Video Title} & \textbf{Category} & \textbf{Content Type} & \textbf{Claims} \\
\midrule
\endhead
\midrule
\multicolumn{4}{r}{Continued on next page} \\
\endfoot
\bottomrule
\endlastfoot
% Climate - Misinformation
Fossil Fuels: The Greenest Energy & Climate & Misinformation & 18 \\
The Great Texas Freeze of 2021 & Climate & Misinformation & 15 \\
Climate Change: What Do Scientists Say? & Climate & Misinformation & 21 \\
Is There Really a Climate Emergency? & Climate & Misinformation & 19 \\
Proof: Worldwide Massive Flooding is All Manmade & Climate & Misinformation & 25 \\
% Climate - Factual
Causes and Effects of Climate Change (NatGeo) & Climate & Factual & 12 \\
Why Does Climate Change Matter & Climate & Factual & 9 \\
Extreme Weather & Climate & Factual & 14 \\
The Life Cycle of a Plastic Bottle & Climate & Factual & 11 \\
What are Greenhouse Gases? & Climate & Factual & 13 \\
\midrule
% Health - Misinformation
Fluoridated Water Lowers IQ (Harvard Study) & Health & Misinformation & 16 \\
Living with HIV: How Women Were Infected & Health & Misinformation & 18 \\
Garlic Cancer & Health & Misinformation & 14 \\
3 Detox Juices & Health & Misinformation & 12 \\
The Magical 3 Day Juice Fast & Health & Misinformation & 15 \\
% Health - Factual
What Happens When You Exercise Regularly & Health & Factual & 17 \\
The Brain-Changing Benefits of Exercise & Health & Factual & 15 \\
Immunology Wars: The Battle with HIV & Health & Factual & 11 \\
Juice vs. Whole Fruit: Which is Healthier? & Health & Factual & 13 \\
What Lack of Sleep Does to the Teenage Brain & Health & Factual & 14 \\
\midrule
% Politics - Misinformation
Egg Price Warning Comes True & Politics & Misinformation & 19 \\
Proof of Election Fraud in 2020 & Politics & Misinformation & 35 \\
FBI Orchestrated Jan 6th & Politics & Misinformation & 22 \\
There is No Gender Wage Gap & Politics & Misinformation & 17 \\
A Nation of Immigrants & Politics & Misinformation & 16 \\
% Politics - Factual
National Trust Sues Trump Admin & Politics & Factual & 20 \\
What Is Democracy (BBC) & Politics & Factual & 9 \\
What is Inflation? & Politics & Factual & 10 \\
UK Election Results Explained & Politics & Factual & 18 \\
Trump's Historic National Emergency & Politics & Factual & 21 \\
\end{longtable}

\normalsize

%-------------------------------------------------------
\subsection{Annotation Schema and Guidelines}
%-------------------------------------------------------

Annotations follow ClaimBuster's check-worthiness principles \citep{hassan2017kdd} and store a compact set of structured fields:

\begin{itemize}
    \item \textbf{Claim text}: Verbatim or lightly paraphrased factual statement taken from the transcript; purely opinionated or rhetorical content is excluded.
    \item \textbf{Importance score}: Thesis-impact hierarchy on a 0.0--1.0 scale:
    \begin{itemize}
        \item 0.85--1.0: Thesis-critical claims whose falsification would collapse the video's main argument
        \item 0.60--0.80: Key evidence claims that significantly support or develop the thesis
        \item 0.30--0.55: Contextual claims that provide background or supplementary information
        \item $<$0.30: Peripheral claims of minimal relevance to the main argument
    \end{itemize}
    \item \textbf{Expected verdict}: Anticipated stance label (SUPPORTS, REFUTES, MIXED, UNCLEAR) with a short rationale grounded in established sources. MIXED applies only when credible evidence exists on both sides, while UNCLEAR denotes insufficient or conflicting evidence under the same criteria across all videos.
\end{itemize}

This schema keeps annotations consistent while capturing the key metadata needed for quantitative evaluation.

%-------------------------------------------------------
\subsection{Dataset Statistics by Category}
%-------------------------------------------------------

\begin{table}[H]
\centering
\caption{Ground truth statistics by category}
\label{tab:gt-stats-category}
\begin{tabular}{lrrrrr}
\toprule
\textbf{Category} & \textbf{Videos} & \textbf{Total Claims} & \textbf{Mean/Video} & \textbf{Factual} & \textbf{Misinfo} \\
\midrule
Climate & 10 & 157 & 15.7 & 59 & 98 \\
Health & 10 & 130 & 13.0 & 70 & 60 \\
Politics & 10 & 216 & 21.6 & 78 & 138 \\
\midrule
\textbf{Total} & \textbf{30} & \textbf{503} & \textbf{16.8} & \textbf{207} & \textbf{296} \\
\bottomrule
\end{tabular}
\end{table}

%-------------------------------------------------------
\subsection{Importance Score Distribution}
%-------------------------------------------------------

\begin{table}[H]
\centering
\caption{Distribution of importance scores across ground truth claims}
\label{tab:importance-distribution}
\begin{tabular}{lrr}
\toprule
\textbf{Importance Range} & \textbf{Count} & \textbf{Percentage} \\
\midrule
High (0.85--1.0) & 127 & 25.2\% \\
Medium-High (0.60--0.84) & 198 & 39.4\% \\
Medium (0.30--0.59) & 143 & 28.4\% \\
Low ($<$0.30) & 35 & 7.0\% \\
\bottomrule
\end{tabular}
\end{table}

%-------------------------------------------------------
\subsection{Expected Verdict Distribution}
%-------------------------------------------------------

\begin{table}[H]
\centering
\caption{Distribution of expected verdicts across ground truth claims}
\label{tab:verdict-distribution}
\begin{tabular}{lrr}
\toprule
\textbf{Expected Verdict} & \textbf{Count} & \textbf{Percentage} \\
\midrule
SUPPORTS & 207 & 41.2\% \\
REFUTES & 189 & 37.6\% \\
MIXED & 68 & 13.5\% \\
UNCLEAR & 39 & 7.7\% \\
\bottomrule
\end{tabular}
\end{table}

%-------------------------------------------------------
\subsection{Data Availability}
%-------------------------------------------------------

The complete annotation dataset, including all claims, importance scores, expected verdicts, and video metadata, is available in the project repository at:

\url{https://github.com/begoechavarren/factible/tree/main/api/experiments/data}

The dataset is provided in YAML format to facilitate both human readability and programmatic access. Each video entry includes:

\begin{itemize}
    \item Video metadata (YouTube ID, title, category, content type)
    \item Complete list of ground truth claims with importance scores and expected verdicts
    \item Tags for filtering and analysis
\end{itemize}
